---
title: "Lecture 4"
author: "Michal Kubi&#353;ta"
date: "2021/10/18"
output:
  xaringan::moon_reader:
    lib_dir: libs
    nature:
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
---

## Structure
- Import & Export
- Sorting
- Aggregation & Mutation
```{r setup, include = FALSE}
library(tidyverse)
```

---
class: middle, inverse

# Import & Export

---
class: middle, inverse

# Sorting

---

## General aproach - order
- can be used for any vector
  - including column of matrix or data.frame
- returns ordering - indexes that can be used to order the data

```{r mtc}
mtc = mtcars %>% select(mpg, cyl, disp, hp) 
head(mtc, 5)
```

---

## Using the ordering
```{r order}
order(mtc$mpg)[1:10]
mtc[order(mtc$mpg),] %>% head(5)
mtc[order(-mtc$mpg, mtc$hp),] %>% head(5)
```

---

## Data.frame specific - arrange

```{r arrange}
mtc %>% 
  arrange(-mpg, hp) %>% 
  head(5)
```

---

class: middle, inverse

# Mutation & Aggregation

---

## Mutation
- adding a new or changing existing columns
```{r mutate}
econ =
    readxl::read_excel("data/econmap.xlsx", 1) %>%
    select(name, year, gdp_cap, energy_productivity, population, labor_force, tfp)

head(econ)
```

---

## Mutation in baseR
- ading new or changing existing columns
```{r mutate_base}
econ$region = ifelse(econ$name == "Afghanistan", "Afghan", "Others") # base R
head(econ)
econ$region = NULL # erase column
colnames(econ)

```

---

## Mutation in Tidyverse
- ading new or changing existing columns

```{r mutate_tidy}
econ %>%
    mutate(region = ifelse(name == "Afghanistan", "Afghan", "Others"),
           tfp = NULL)
```

```{r region, include = FALSE}
econ =
    econ %>%
    mutate(region = ifelse(name == "Afghanistan", "Afghan", "Others"),
           tfp = NULL)
```

---

## Aggregation
- calculate function by group

```{r aggragate}
econ %>% 
    group_by(name) %>% 
    summarise(gdp = mean(gdp_cap),
              energy = sum(energy_productivity)
              )
```

---

## Aggregation on multi-index

```{r aggregate_multi}
econ %>% 
    group_by(region, year) %>% 
    summarise(pop = mean(population),
              labor = mean(labor_force)) %>% 
    ungroup() %>% 
    mutate(active = labor / pop) 
```

---

## Aggregation with multi-value output
- functions can return several values for each group
- typically used for top X items per group 

```{r aggregate_top}
econ %>% 
  group_by(year) %>% 
  top_n(3, gdp_cap) %>% 
  arrange(year, -gdp_cap) %>% 
  head(10)
```